An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects\\ in Time-Frequency Representations

### 2020-07-24 Notes/BoundEffRed.tex
notes

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the real-time reduction of boundary effects in TF representations. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient to extend signals formed by a sum of sine waves. Moreover, thanks to the low computational time, the numerical results show that this strategy allows us to switch to a real-time implementation of {\sf BoundEffRed}, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many TF representations. 
%Other numerical examples are also included, such as the application of {\sf BoundEffRed} to other types of biomedical signals such as ECGs.

Various improvements can be considered to make the algorithm more robust. In particular, we have noticed (see Fig.~\ref{fig:THO.failure}) that when the regular oscillations of the observed signal break, the forecasting step is no longer relevant, and only slows down the calculation of the TF representation. A preliminary step should then be added to the algorithm to detect signal activity, and disable the forecasting step when possible. A more fundamental improvement would be to perform the forecasting step directly in the TF plane, to specifically predict the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series, with each frequency band containing a univariate time series. We will explore these possibilities in our future work.

>>>>

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the real-time reduction of boundary effects in TF representations. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient to extend signals formed by a sum of sine waves. Moreover, the low computational time allows us to switch to a real-time implementation of {\sf BoundEffRed}, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many TF representations. An additional application to ECG is included into the supplementary material.

Various improvements can be considered to make the algorithm more robust. In particular, we have noticed (see Fig.~\ref{fig:THO.failure}) that when the regular oscillations of the observed signal break, the forecasting step is no longer relevant, and only slows down the calculation of the TF representation. A preliminary step should then be added to the algorithm to detect signal activity, and disable the forecasting step when possible. A more fundamental improvement would be to perform the forecasting step directly in the TF plane, to specifically predict the evolution of each instantaneous frequency. This leads to seeing a TF representation as a multivariate time series, with each frequency band containing a univariate time series. These avenues will be explored in our future work.

### 2020-07-24 Notes/Supplementary material/SupMat.tex
small updates
paper
notes

\usepackage{graphicx}
\usepackage{bbm}
\usepackage{amsfonts,amssymb,amsmath,amsthm}
\usepackage{type1cm,eso-pic,color}
\usepackage{mathrsfs} 
\usepackage{mathpazo}
\usepackage[scaled=.95]{helvet}
\usepackage{courier}
\usepackage{xspace}
\usepackage{etoolbox}
\usepackage[shortlabels]{enumitem}
\usepackage{multirow}
\usepackage{dsfont}

@@

\title{An Approach for the Reduction of Boundary Effects in Time-Frequency Representations:\\ Part II}

@@

\setcounter{equation}{26}

@@

It follows from this:
\begin{align*}
\tilde\bA &= (\bS^{(1)}+\bE^{(1)})(\bS^{(0)}+\bE^{(0)})^\inv \\
\bA_0 &=\bS^{(1)}{\bS^{(0)}}^\inv\ .
\end{align*}
Then:
\begin{align}
\nonumber
\bh^{(\ell)} &= \balpha^{(\ell)} - \balpha^{(\ell)}_0 \\
\nonumber
&= \be_M^T\left(\tilde\bA^\ell-\tilde\bA_0^\ell\right) \\
&= \be_M^T\left(\left((\bS^{(1)}+\bE^{(1)})(\bS^{(0)}+\bE^{(0)})^\inv\right)^\ell- \tilde\bA_0^\ell\right)\ .
\label{eq:error.vec}
\end{align}

\subsection{Study of $\bh^{(\ell)}$}
The randomness of $\bh^{(\ell)}$ is completely originating from the error matrices. Besides, notice that the first $M-1$ rows in $\bE^{(1)}$ are equals to the last $M-1$ rows of $\bE^{(0)}$. Thus, we gather all the sources of randomness into an vector $\bg\in\RR^{M(M+1)}$, containing $M$ rows and defined as:
\[
\bg = \vec\left(
\begin{pmatrix}
\bE^{(0)} \\
\be_M^T\bE^{(1)}
\end{pmatrix}
\right)\ .
\]
Here, "$\vec$" denotes the vectorization operator, concatenating the columns of a given matrix on top of one another. Then, by definition, we have:
\[
\bg = \sigma\bg_1 + \sigma^2\bg_2\ ,
\]
where:
\begin{align*}
\bg_1 &= \dfrac1K\sum_{k=0}^{K-1}\vec\left( \tilde\bz_k\bw_k^T + \tilde\bw_k\bz_k^T \right) \\
\bg_2 &= \dfrac1K\sum_{k=0}^{K-1}\vec\left(\tilde\bw_k\bw_k^T-\tilde\bI \right) \ .
\end{align*}
with $\tilde\bz_k^T= \begin{pmatrix} \bz_k^T & \bz_{k+1}[0] \end{pmatrix}$ and $\tilde\bw_k^T= \begin{pmatrix} \bw_k^T & \bw_{k+1}[0] \end{pmatrix}$. Then, $\bg_1$ is a Gaussian random vector because it is a linear combination of Gaussian random vectors. Moreover, using the central limit theorem under weak dependence, we can show that $\bg_2$ also converges towards a Gaussian random vector as $K\to\infty$. Combining these two results gives the following result:
\[
\sqrt{K}\ \bg \xrightarrow[K\to\infty]{\cD} \cN(\bzero,\bGamma_0)\ ,
\]
where $\bGamma_0=\EE\left\{\bg\bg^T\right\}$ is a covariance matrix.

@@

\subsection{Expression of the bias $\bmu$.}

@@

\subsection{Expression of the covariance $\bgamma$.}
We recall that $n>N$, we consequently denote $n=N-1+\ell$.  To derive the covariance, let us segregate the cases.

@@

%\begin{align*}
%\bgamma[\ell,\ell] &\leq \dfrac1K\left(\sqrt{\bz_K^T\EE\{{\bu^{(\ell)}}^T\bu^{(\ell)}\}\bz_K} + \sigma\sqrt{\EE\{(\bu^{(\ell)}\bw_K)^2\}}\right)^2 \\
%&\ + \dfrac{2\sigma}{\sqrt{K}}\|\balpha_0^{(\ell)}\|\left(\sqrt{\bz_K^T\EE\{{\bu^{(\ell)}}^T\bu^{(\ell)}\}\bz_K} +\sigma\sqrt{\EE\{(\bu^{(\ell)}\bw_K)^2\}} \right)+ \sigma^2\|\balpha_0^{(\ell)}\|^2 - \bmu[\ell]^2 + o(\sigma^2) \\
%\end{align*}
%
%Then, Isserlis formula gives us:
%\begin{align*}
%\EE\{(\bu^{(\ell)}\bw_K)^2\} &= \sum_{m,m'=0}^{M-1}\EE\{\bu^{(\ell)}[m]\bw_K[m]\bu^{(\ell)}[m']\bw_K[m']\}\\
%&\underset{K\to\infty}{\sim} \sum_{m,m'=0}^{M-1}\EE\{\bu^{(\ell)}[m]\bw_K[m]\}\EE\{\bu^{(\ell)}[m']\bw_K[m']\}+\EE\{\bu^{(\ell)}[m]\bu^{(\ell)}[m']\}\EE\{\bw_K[m]\bw_K[m']\} \\
%&\hspace{30pt}+\EE\{\bu^{(\ell)}[m]\bw_K[m']\}\EE\{\bu^{(\ell)}[m']\bw_K[m]\}\\
%&\leq \Tr(\bGamma^{(\ell)}) + 2\sum_{m,m'=0}^{M-1}\sqrt{\bGamma^{(\ell)}[m,m]\bGamma^{(\ell)}[m',m']}
%\end{align*}
%So:
%\begin{align*}
%\bgamma[\ell,\ell] &\leq \dfrac1K\left(\sqrt{\bz_K^T\bGamma^{(\ell)}\bz_K} + \sigma\sqrt{\Tr(\bGamma^{(\ell)}) + 2\sum_{m,m'=0}^{M-1}\sqrt{\bGamma^{(\ell)}[m,m]\bGamma^{(\ell)}[m',m']}}\right)^2 \\
%&\ + \dfrac{2\sigma}{\sqrt{K}}\|\balpha_0^{(\ell)}\|\left(\sqrt{\bz_K^T\bGamma^{(\ell)}\bz_K} +\sigma\sqrt{\Tr(\bGamma^{(\ell)}) + 2\sum_{m,m'=0}^{M-1}\sqrt{\bGamma^{(\ell)}[m,m]\bGamma^{(\ell)}[m',m']}} \right)+ \sigma^2\|\balpha_0^{(\ell)}\|^2 + o(\sigma^2) \\
%\end{align*}

>>>>

\usepackage{graphicx}
\usepackage{bbm}
\usepackage{amsfonts,amssymb,amsmath,amsthm}
\usepackage{type1cm,eso-pic,color}
\usepackage{mathrsfs} 
\usepackage{mathpazo}
\usepackage[scaled=.95]{helvet}
\usepackage{courier}
\usepackage{xspace}
\usepackage{etoolbox}
\usepackage[shortlabels]{enumitem}
\usepackage{multirow}
\usepackage{dsfont}
\usepackage{hhline}
\usepackage{graphicx}

@@

\title{An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects\\ in Time-Frequency Representations:\\ Part II}

@@

\setcounter{equation}{27}
\setcounter{figure}{7}
\setcounter{table}{3}

@@

It follows from this:
\begin{align*}
\tilde\bA &= (\bS^{(1)}+\bE^{(1)})(\bS^{(0)}+\bE^{(0)})^\inv \\
\bA_0 &=\bS^{(1)}{\bS^{(0)}}^\inv\ .
\end{align*}
Then:
\begin{align}
\nonumber
\bh^{(\ell)} &= \balpha^{(\ell)} - \balpha^{(\ell)}_0 \\
\nonumber
&= \be_M^T\left(\tilde\bA^\ell-\bA_0^\ell\right) \\
&= \be_M^T\left(\left((\bS^{(1)}+\bE^{(1)})(\bS^{(0)}+\bE^{(0)})^\inv\right)^\ell- \bA_0^\ell\right)\ .
\label{eq:error.vec}
\end{align}

\subsection{Study of $\bh^{(\ell)}$}
The randomness of $\bh^{(\ell)}$ is completely originating from the error matrices. Besides, notice that the first $M-1$ rows in $\bE^{(1)}$ are equals to the last $M-1$ rows of $\bE^{(0)}$. Thus, we gather all the sources of randomness into an vector $\bg\in\RR^{M(M+1)}$, containing $M$ rows and defined as:
\[
\bg = \vec\left(
\begin{pmatrix}
\bE^{(0)} \\
\be_M^T\bE^{(1)}
\end{pmatrix}
\right)\ .
\]
Here, "$\vec$" denotes the vectorization operator, concatenating the columns of a given matrix on top of one another. Then, by definition, we have:
\[
\bg = \sigma\bg_1 + \sigma^2\bg_2\ ,
\]
where:
\begin{align*}
\bg_1 &= \dfrac1K\sum_{k=0}^{K-1}\vec\left( \tilde\bz_k\bw_k^T + \tilde\bw_k\bz_k^T \right) \\
\bg_2 &= \dfrac1K\sum_{k=0}^{K-1}\vec\left(\tilde\bw_k\bw_k^T-\tilde\bI \right) \ .
\end{align*}
with $\tilde\bz_k^T= \begin{pmatrix} \bz_k^T & \bz_{k+1}[M-1] \end{pmatrix}$ and $\tilde\bw_k^T= \begin{pmatrix} \bw_k^T & \bw_{k+1}[M-1] \end{pmatrix}$. Then, $\bg_1$ is a Gaussian random vector because it is a linear combination of Gaussian random vectors. Moreover, using the central limit theorem under weak dependence, we can show that $\bg_2$ also converges towards a Gaussian random vector as $K\to\infty$. Combining these two results gives the following result:
\[
\sqrt{K}\ \bg \xrightarrow[K\to\infty]{\cD} \cN(\bzero,\bGamma_0)\ ,
\]
where $\bGamma_0=\EE\left\{\bg\bg^T\right\}$ is a covariance matrix.

@@

\subsection{Expression of the Bias $\bmu$.}

@@

\subsection{Expression of the Covariance $\bgamma$.}
We recall that $n>N$, we consequently denote $n=N-1+\ell$.  To derive the covariance, let us segregate the cases.

@@

\section{Application to an Electrocardiogram}
We provide here an additional implementation of {\sf BoundEffRed}, applied to an electrocardiogram (ECG) dataset. The dataset is constructed from a $500$-second-long ECG, sampled at $\fs=200$~Hz, cut into $14$ segments of $35$ seconds each. Fig.~\ref{fig:ecg} depicts one of these subsignals, together with the estimated 5 seconds extensions obtained by {\sf SigExt}, EDMD, and GPR forecastings  are superimposed to the ground-truth extension. The sharp and spiky ECG patterns make the AHM model too simplistic to describe this type of signal. Consequently, the forecast produced by {\sf SigExt} is moderately satisfactory.

\begin{figure}
\includegraphics[width=\textwidth]{ECGforecast.eps}
\caption{Extended ECG (blue) obtained by the {\sf SigExt} forecasting (top), the EDMD forecasting (middle), and the GPR forecasting (bottom), superimposed with the ground truth signal (red).}
\label{fig:ecg}
\end{figure}

Table~\ref{tab:otd.ecg} contains the performance index $D$ of the boundary-free TF representations, averaged over the $N$ subsignals, according to the extension method. As a result of the fair quality of the forecasts, the reduction of boundary effects is less significant than for PPG signal. Nevertheles, the results show that {\sf BoundEffRed} has the same efficiency when the {\sf SigExt} extension, the EDMD extension or the GPR extension is chosen. This justifies the choice of {\sf SigExt} for real-time implementation.

\begin{table}
\centering
\caption{ECG: Performance of the Boundary-Free TF Representations According to the Extension Method}
\begin{tabular}{|c||c|c|c|c|}
  \hline
   \multirow{2}{40pt}{\centering Extension method} & \multicolumn{4}{c|}{Averaged performance index $D$} \\
   \cline{2-5}
      & STFT & SST & ConceFT & RS\\
   \hhline{|=#=|=|=|=|}
   {\sf SigExt} & $0.713$ & $0.700$ & $0.753$ & $0.761$ \\
   \hline
   Symmetric & $1.395$ & $1.402$ & $1.416$ & $0.976$ \\
   \hline
   EDMD & $0.856$ & $0.792$ & $0.559$ & $0.688$ \\
   \hline
   GPR & $0.723$ & $0.734$ & $0.734$ & $0.761$ \\
   \hline
\end{tabular}
\label{tab:otd.ecg}
\end{table}

### 2020-07-24 Notes/results.tex
notes

We then apply {\sf BoundEffRed} (\ie~Algorithm~\ref{alg:boundary}) for diverse TF representations: STFT, SST, RS, as well as \textit{Concentration of Frequency and Time} (ConceFT), a generalized multitaper SST-based representation introduced in~\cite{Daubechies16conceft}. In this application and throughout the document, we use a Gaussian window as the analysis window, regardless of the representation involved. In Table~\ref{tab:THO}, we give the averaged performance index~\eqref{eq:index.perf}, evaluated the whole TF representations (including boundaries). Even though {\sf SigExt} performs somehow moderately, the boundary effects are dramatically reduced on the TF representations, in the same order of magnitude than with the forecastings given by EDMD or GPR. Notice that the extension length $L$ has been set accordingly to the window length used by the TF analysis tool. For instance, the window length used to evaluate the STFT is of $1500$ samples. To prevent the STFT from being sensitive to the boundaries, we set $L=750$. In this way, the evaluation of the spectral content of the signal near its boundaries is not limited by a lack of information all along the window support. From now on, all results are given for $L$ equal to the half of the width of the window used in the TF transform.

@@

We divide the signal into 32-second-long segments and apply Algorithm~\ref{alg:boundary} on each piece. We provide in Table~\ref{tab:otd.ppg} the performance index $D$ of the boundary-free TF representations averaged over the signals. For all the considered TF representations, the results clearly show that our algorithm reduces the boundary effects. Even though on this signal, the {\sf SigExt} extension yields TF representations slightly more sensitive to boundary effects than the extensions given by EDMD or GPR, it is the only technique that allows a real-time implementation.

>>>>

We then apply {\sf BoundEffRed} for diverse TF representations: STFT, SST, RS, as well as \textit{Concentration of Frequency and Time} (ConceFT), a generalized multitaper SST-based representation introduced in~\cite{Daubechies16conceft}. In this application and throughout the document, we use a Gaussian window as the analysis window, regardless of the representation involved. In Table~\ref{tab:THO}, we give the averaged performance index~\eqref{eq:index.perf}, evaluated the whole TF representations (including boundaries). Even though {\sf SigExt} performs somehow moderately, the boundary effects are dramatically reduced on the TF representations, in the same order of magnitude than with the forecastings given by EDMD or GPR. Notice that the extension length $L$ has been set accordingly to the window length used by the TF analysis tool. For instance, the window length used to evaluate the STFT is of $1500$ samples. To prevent the STFT from being sensitive to the boundaries, we set $L=750$. In this way, the evaluation of the spectral content of the signal near its boundaries is not limited by a lack of information all along the window support. From now on, all results are given for $L$ equal to the half of the width of the window used in the TF transform.

@@

We divide the signal into 32-second-long segments and apply {\sf BoundEffRed} on each piece. We provide in Table~\ref{tab:otd.ppg} the performance index $D$ of the boundary-free TF representations averaged over the signals. For all the considered TF representations, the results clearly show that our algorithm reduces the boundary effects. In addition to being the only technique that allows real-time implementation, {\sf SigExt} produces TF representations that erase boundary effects about as efficiently as extensions given by EDMD or GPR.
